---
title: "Spatial Microsimulation for Middlesex County, NJ"
author: "Arpan Nookala"
date: "2024-06-23"
output: html_document
format: 
   html:
     df-print: paged
---

```{r}
# Loading Required Libraries
library(dplyr)
library(tidyr)
```

## Reading the Individual Level Data

```{r}
# Read the Household Pulse Survey data from a CSV file
ind <- read.csv("hps_04_00_01_puf.csv")
# Display the data
head(as.data.frame(ind))
```

## Data Filtering and Variable Selection

```{r}
# Filter data for Middlesex County (EST_MSA=35620) and state of New Jersey (EST_ST=34)
ind <- ind %>% filter(EST_MSA==35620 & EST_ST==34)

# Calculate Age from Birth Year
ind$AGE <- 2024 - ind$TBIRTH_YEAR

# Select variables of interest
variables_of_interest <- c("EST_ST","EST_MSA", "MS", "RRACE", "TWDAYS", "EEDUC", "KINDWORK", "THHLD_NUMPER", "INCOME","EGENID_BIRTH","AGE","RHISPANIC","ANYWORK")
ind <- ind %>% select(all_of(variables_of_interest))

# Display the filtered and selected data
head(as.data.frame(ind))
```

## Data Cleaning and Transformations

```{r}
# Display unique values of variables for understanding
sort(unique(ind$AGE))
sort(unique(ind$EEDUC))
sort(unique(ind$EGENID_BIRTH))
sort(unique(ind$KINDWORK))

# Replace special values with NA in education variable
ind$EEDUC[ind$EEDUC == -99] <- NA
ind$EEDUC[ind$EEDUC == -88] <- NA

# Drop rows with NA values
ind <- ind %>% drop_na()

# Filter data for individuals aged 25 and above
ind <- ind[ind$AGE >= 25,]

# Display cleaned data
head(as.data.frame(ind))

# Display unique values again after cleaning
sort(unique(ind$AGE))
sort(unique(ind$EEDUC))
sort(unique(ind$EGENID_BIRTH))
sort(unique(ind$KINDWORK))
```

## Recoding Variables
```{r}
# Define age group breaks and labels
brks <- c(25, 30, 35, 40, 45, 50, 55, 60, 62, 65, 67, 70, 75, 80, 85, Inf)
labs <- c("25 to 29 years", "30 to 34 years", "35 to 39 years", "40 to 44 years", 
          "45 to 49 years", "50 to 54 years", "55 to 59 years", "60 and 61 years", 
          "62 to 64 years", "65 and 66 years", "67 to 69 years", "70 to 74 years", 
          "75 to 79 years", "80 to 84 years","85 years and over")

# Recode age into groups
ind$AGE <- cut(ind$AGE, breaks = brks, labels = labs, right=FALSE)

# Recode gender variable
ind$EGENID_BIRTH <- factor(ind$EGENID_BIRTH, levels = 1:2, labels = c("Male", "Female"))

# Recode education levels
ind$EEDUC <- factor(ind$EEDUC, levels = 1:7, labels = c("Less than high school","Some high school","High school graduate or equivalent (for example GED)","Some college, but degree not received or is in progress","Associate’s degree (for example AA, AS)","Bachelor's degree (for example BA, BS, AB)","Graduate degree (for example master's, professional, doctorate)"))

# Display recoded data
head(as.data.frame(ind))
```

## Reading and Validating Constraint Data
```{r}
# Read constraint data for age, sex, and education from CSV files
con_age <- read.csv("con_age.csv")
con_sex <- read.csv("con_sex.csv")
con_edu <- read.csv("con_edu.csv")

# Display the constraint data
head(as.data.frame(con_age))
head(as.data.frame(con_sex))
head(as.data.frame(con_edu))

# Validate the sums of the constraints
# Sum of age constraints
sum(con_age)
# Sum of sex constraints
sum(con_sex)
# Sum of education constraints
sum(con_edu)

# Validate the row sums of the constraints
# Row sums of age constraints
rowSums(con_age)
# Row sums of sex constraints
rowSums(con_sex)
# Row sums of education constraints
rowSums(con_edu)

# Check if the row sums of different constraints are equal
# Check if row sums of age and sex constraints are equal
rowSums(con_age) == rowSums(con_sex)
# Check if row sums of age and education constraints are equal
rowSums(con_age) == rowSums(con_edu)
# Check if row sums of sex and education constraints are equal
rowSums(con_sex) == rowSums(con_edu)
```

## Naming Columns and Combining Constraints
```{r}
# Assign column names to age and education constraints based on the levels in the individual data
names(con_age) <- levels(ind$AGE)
names(con_edu) <- levels(ind$EEDUC)

# Combine age, sex, and education constraints into a single data frame
cons <- cbind(con_age, con_sex, con_edu)

# Display the combined constraints
head(as.data.frame(cons))
```

## Creating Categorical Indicator Matrices
```{r}
# Create binary indicator matrix for age categories
cat_age <- model.matrix(~ ind$AGE - 1)
# Display the dimensions of the age indicator matrix
dim(cat_age)

# Create binary indicator matrix for sex categories
cat_sex <- model.matrix(~ ind$EGENID_BIRTH - 1)
# Display the dimensions of the sex indicator matrix
dim(cat_sex)

# Create binary indicator matrix for education categories
cat_edu <- model.matrix(~ ind$EEDUC - 1)
# Display the dimensions of the education indicator matrix
dim(cat_edu)

# Combine age, sex, and education indicator matrices into a single matrix
ind_cat <- cbind(cat_age, cat_sex, cat_edu)

# Display the first few rows of the combined indicator matrix
head(as.data.frame(ind_cat))
```

## Aggregating Indicator Matrix and Defining Dimensions
```{r}
# Calculate the column sums of the indicator matrix
colSums(ind_cat)

# Aggregate the individual categories
ind_agg <- colSums(ind_cat)

# Compare the first row of constraints with the aggregated individual categories
rbind(cons[1,], ind_agg)

# Define the number of zones (rows) in the constraints data
n_zone <- nrow(cons)
# Display the number of zones
n_zone

# Define the number of individuals in the individual data
n_ind <- nrow(ind)
# Display the number of individuals
n_ind

# Define the number of age categories
n_age <- ncol(con_age)
# Display the number of age categories
n_age

# Define the number of sex categories
n_sex <- ncol(con_sex)
# Display the number of sex categories
n_sex

# Define the number of education categories
n_edu <- ncol(con_edu)
# Display the number of education categories
n_edu
```

## Initializing Weight Matrices and Performing First Iteration of IPF
```{r}
# Initialize weight matrices with uniform weights
weights <- matrix(data = 1, nrow = nrow(ind), ncol = nrow(cons))
weights3 <- weights1 <- weights2 <- weights
# Display the dimensions of the weights matrix
dim(weights)

# Aggregate the individual categories initially, based on constraints
ind_agg0 <- t(apply(cons, 1, function(x) 1 * ind_agg))
colnames(ind_agg0) <- names(cons)
# Display the initial aggregated individual categories
head(as.data.frame(ind_agg0))

# First iteration of IPF to adjust weights based on age constraints
for(j in 1:n_zone){
  for(i in 1:n_age){
    index <- ind_cat[, i] == 1
    weights1[index, j] <- weights[index, j] * con_age[j, i] / ind_agg0[j, i]
  }
}
# Display the adjusted weights after the first iteration
head(as.data.frame(weights1))
```

## Aggregating After First IPF Iteration
```{r}
# Initialize aggregated matrices to store results of each IPF iteration
ind_agg3 <- ind_agg2 <- ind_agg1 <- ind_agg0 * NA

# Aggregate individual categories after the first IPF iteration
for(i in 1:n_zone){
  ind_agg1[i, ] <- colSums(ind_cat * weights1[, i])
}
# Display the aggregated individual categories after the first iteration
head(as.data.frame(ind_agg1))

# Calculate the row sums of the aggregated categories and the constraints for age
rowSums(ind_agg1[, 1:15])
rowSums(cons[, 1:15])

# Check if the row sums of the aggregated categories match the constraints for age
rowSums(ind_agg1[, 1:15]) == rowSums(cons[, 1:15])

```

## Second Iteration of IPF and Validation
```{r}
# Second iteration of IPF to adjust weights based on sex and age constraints
for (j in 1:n_zone) {
  for (i in 1:(n_sex + n_age)) {
    index <- ind_cat[, i] == 1
    weights2[index, j] <- weights1[index, j] * cons[j, i] / ind_agg1[j, i]
  }
}
# Display the adjusted weights after the second iteration
head(as.data.frame(weights2))

# Aggregate individual categories after the second IPF iteration
for(i in 1:n_zone){
  ind_agg2[i, ] <- colSums(ind_cat * weights2[, i])
}
# Display the aggregated individual categories after the second iteration
head(as.data.frame(ind_agg2))

# Calculate and display the row sums of the aggregated categories for sex
rowSums(ind_agg2[, 16:17])

# Calculate and display the row sums of the constraints for sex
rowSums(cons[, 16:17])

# Check if the row sums of the aggregated categories match the constraints for sex
rowSums(ind_agg2[, 16:17]) == rowSums(cons[, 16:17])
```

## Third Iteration of IPF and Final Validation
```{r}
# Third iteration of IPF to adjust weights based on sex, age, and education constraints
for(j in 1:n_zone){
  for(i in 1:(n_sex + n_age + n_edu)){
    index <- ind_cat[, i] == 1
    if(ind_agg2[j, i] != 0) {  # Check to avoid division by zero
      weights3[index, j] <- weights2[index, j] * cons[j, i] / ind_agg2[j, i]
    }
  }
}
# Display the adjusted weights after the third iteration
head(as.data.frame(weights3))

# Aggregate individual categories after the third IPF iteration
for(i in 1:n_zone){
  ind_agg3[i, ] <- colSums(ind_cat * weights3[, i])
}
# Display the aggregated individual categories after the third iteration
head(as.data.frame(ind_agg3))

# Calculate and display the row sums of the aggregated categories for education
rowSums(ind_agg3[, 18:24])

# Calculate and display the row sums of the constraints for education
rowSums(cons[, 18:24])

# Check if the row sums of the aggregated categories match the constraints for education
rowSums(ind_agg3[, 18:24]) == rowSums(cons[, 18:24])
```

## Correlation Analysis Between Aggregated Data and Constraints
```{r}
# Function to convert matrix to numeric vector
vec <- function(x) as.numeric(as.matrix(x))

# Calculate and display the correlation between the initial aggregated data and constraints
cor(vec(ind_agg0), vec(cons))

# Calculate and display the correlation between the first iteration aggregated data and constraints
cor(vec(ind_agg1), vec(cons))

# Calculate and display the correlation between the second iteration aggregated data and constraints
cor(vec(ind_agg2), vec(cons))

# Calculate and display the correlation between the third iteration aggregated data and constraints
cor(vec(ind_agg3), vec(cons))
```